---
title: Week 6 Activity 1
subtitle: Use of Color in Graphs 
format: html
---
## Introduction
This activity gives you practice applying the concepts we just learned about color mapping and scales in data visualization. You will recreate several plots while learning how to control the appearance of colors using `scale_*()` functions in **ggplot2**. You will work with functions that map variables to visual properties and customize color palettes in your plots.

:::::: columns
::: {.column}
**Core ggplot functions** \

- `ggplot()` \ 

- `aes()` \

- `geom_point()` \

- `geom_bar()` \

- `geom_col()` \

- `color` \ 

- `fill` \

:::

::: {.column}
**Scale & customization functions**
- `scale_color_manual()` \ 

- `scale_color_gradient()` \ 

- `scale_color_continuous()` \

- `scale_fill_manual()` \

- `scale_fill_discrete()` \

- `labs()` \

- `theme_*()` \
:::
::::::

This activity is intended to be finished during classtime. In the case that we run out of time, please complete it indepedently and submit it via BCourses by tonight (Tuesday) at 11:59PM. To submit to BCourses, \

- *Render your QMD* (Render button in RStudio and Preview button in Positron) \ 

- Your HTML file will be created and can be found in the same folder on your computer that the QMD is stored in. \ 

- Submit both the HTML and QMD files to the 'Week 6 Activity' assignment in BCourses. \

*If any of these functions are unfamiliar to you, you can access the help documentation by typing `?function_name` into your R console. Alternatively, you can review R basics in the following [textbook](https://www.gastonsanchez.com/R-coding-basics/).*

**1.  Load the required packages and demo code.**

```{r}
#| eval: false
#| output: false
library(ggplot2)
library(viridis)

set.seed(123)

demo_df <- data.frame(
  x = rnorm(20),
  y = rnorm(20),
  logical_var = sample(c(TRUE, FALSE), 20, replace = TRUE),
  integer_var = sample(1:5, 20, replace = TRUE),
  double_var = runif(20, 0, 10),
  factor_var = factor(sample(letters[1:3], 20, replace = TRUE)),
  char_var = sample(c("cat", "dog", "bird"), 20, replace = TRUE)
)

head(demo_df)
```

*Hint: if you have not yet installed one of these packages, you must first run `install.packages("package_name")` in your R console.*

## Part 1: Play around with using functions to control colors 

**2. Recreate the scatterplot so that each animal category has a specific color.**
Target colors \

- cat → orange \

- dog → steelblue \

- bird → forestgreen \

```{r}
#| eval: false
ggplot(demo_df, aes(x, y, color = char_var)) +
  geom_point(size = 4) +
  __________(
    values = c(
      "cat" = ________,
      "dog" = ________,
      "bird" = ________
    )
  ) +
  labs(title = "Custom Discrete Colors")
```

*Hint: Try all options: \

- `scale_color_manual()` \

- `scale_color_continuous()` \

- `scale_fill_manual()`.* \

**3. Recreate the scatterplot using a color gradient that represents the numeric variable `double_var`.** 
Use:
- lightblue for low values \

- darkblue for high values

```{r} 
#| eval: false
ggplot(demo_df, aes(x, y, color = double_var)) +
  geom_point(size = 4) +
  __________(
    low = "________",
    high = "________"
  ) +
  labs(title = "Custom Continuous Scale")
```

*Hint: Try all options: \

- `scale_color_manual()` \

- `scale_color_gradient()` \

- `scale_fill_discrete()`.* \

**4. Recreate the bar plot using filled colors for each category.**
```{r}
#| eval: false
ggplot(demo_df, aes(x = factor_var, fill = char_var)) +
  geom_bar() +
  __________() +
  labs(title = "Bar Plot with Discrete Fill")
```

*Hint: Try all options: \

- `scale_color_discrete()` \

- `scale_fill_discrete()` \

- `scale_color_gradient()`.*