---
title: Week 10 Activity 1
subtitle: NYT Graph Recreation
format: html
---

# Data 

Results of the 2020 Presidential Election in the U.S. are available in the *Electoral College* website:

<https://www.archives.gov/electoral-college/2020>

Specifically, we can find data of *Electoral College Certificates and Votes by State*.

As you can tell from the link above, the data is in an HTML table. Because of this, we can import the data in R with the `read_html()` and `html_table()` functions.

```{r}
url = "https://www.archives.gov/electoral-college/2020"

tbls = read_html(url) |>
  html_table()

# we want 2nd table (i.e. table of electoral votes)
tbl = tbls[[2]]
```

Recall that `html_table()` extracts ALL the html tables in an html document. In this case, the data happens to be in the 2nd table.

## Data Preparation

The data in `tbl` is not quite ready to be crunched. We need to do some adjustments:

```{r}
# renaming columns
names(tbl) = 

# getting rid of 1st and last 2 rows
tbl = tbl[-c(1, 53:54), ]

tail(tbl)
```

To keep things simple, let's focus only on columns `State`, `Pres_Dem`, and `Pres_Rep`. We'll subset these columns into their own table `dat`. We also need to take care of some strings

```{r}
dat = select(tbl, c(1, 3, 4))

# replacing dashes to zeroes (and making them numeric)


# getting rid of asterisks in State names

```

```{r}
# clean table
dat = dat |>
  

head(dat)
```

------------------------------------------------------------------------

# Graphs: Choropleth Maps

Let's go over a series of plotting rounds, starting with a basic map, and then gradually adding more elements, and customizing its appearance to get as close as possible to our target visualization.

## Map of USA from `"rnaturalearth"`

We begin by producing a first map of the U.S. This requires to get map data of the US territory---at the State level---via `ne_states()`. Notice that the `returnclass` argument must be set as `returnclass = "sf"` to guarantee that the return map object is of class `"sf"` (simple features).

With the `"sf'` object in place, we can pass it to `ggplot()` and its layer function `geom_sf()`

Run the following cell ONCE by removing all the '#'s' and then putting them back or deleting the cell

```{r}
# remotes::install_github("ropensci/rnaturalearthhires")

# install.packages(
  #"rnaturalearthhires",
  #repos = "https://ropensci.r-universe.dev",
  #type = "source"
  #)

```

```{r}
usa_map = 

ggplot() + 
  geom_sf(data = usa_map)
```

As you can tell, we are able to produce a map of the U.S. but it's not the pretiest map.

## Map of Contiguous States

Let's focus on contiguous States only by adjusting the latitude (x-axis) and longitude (y-axis) coordinates with `coord_sf()`. We can also simplify the theme with `theme_void()`

```{r}
ggplot() + 
  
```

## Merging map-data with elections-data

So far we have a map of the US States. But what about the elections data? We need to merge (join) the table of the map `usa_map` with the table of the 2020 election results `dat`. This merging operation is done with `inner_join()`, indicating the names of the columns in both tables that have the States: `name` from table `usa_map`, and `State` from table `dat`

```{r}
# merging map and votes data tables
usa_map_2020 = usa_map |>
  
```

With the merged data `usa_map_2020`, we can plot again a map of the US contiguous states:

```{r}
# US contiguous states
ggplot(data = usa_map_2020) + 
  
```

Obviously this map is no different from the preceding map. This is just a sanity check to make sure the map data has not been corrupted.

## Coloring the Map

What characterizes a choropleth map is the use of one or more variables to color code the polygons or areas in the desired map.

We want to color code the states according to the party of the winner candidate. This information is in the variable (or column) `Party` which we can map to the `fill` argument inside `geom_sf()`:

```{r}
# 1st version
ggplot(data = usa_map_2020) + 
  
```

## Choropleth Map 2

As you may know, the color of the Democratic party is blue, whereas the color of the Republican party is red. To change the `fill` colors we can set new values with `scale_fill_manual()`

```{r}
# 2nd version: customize colors
ggplot(data = usa_map_2020) + 
  
```

## Choropleth Map 3

If you look at the choropleth map published in the NYT, the color hues that they use are less saturated than the default `blue` and `red` colors in R.

We can inspect the HTML code to see the exact color values (in hexadecimal notation) used by the NYT.

- blue: `"#1375b7"`

- red: `"#c93135"`

Let's update the color scale in our choropleth map:

```{r}
# 3rd version
ggplot(data = usa_map_2020) + 
  
```

## Choropleth Map 4

The 2-letter abbreviation of the State names are in column `postal`; we can use these to add the state names on the map.

```{r}
usa_map_2020$postal
```

```{r}
# 4th version
ggplot(data = usa_map_2020) + 
  
```

## Choropleth Map 5

The last tweak involves coloring the text of the State abbreviations:

```{r}
# 5th Attempt
ggplot(data = usa_map_2020) + 
  
```