---
title: Week 3
subtitle: Data Types 
format: stat198slides-revealjs
---

# Introduction

This is a sample deck for DataViz DeCal Slides. The features include:

1. Title slide with background image.
2. A bunch of yaml defaults.
3. Clickable logo in lower right that links to the previous page.
4. No logo or slide number on title slide.
5. Custom `:::poll` div for poll questions.
6. A [countdown timer](https://github.com/gadenbuie/countdown/tree/main/quarto) accessible by adding

   `{{< countdown "01:30" >}}`
   
   on a new line.


# Agenda

- Discussion of Assigned Reading   
- Lecture on understanding different types of variables and how they influence chart selection. 
- Activity 
- Attendence quiz 

# Discussion of Assigned Reading 

- **Reading:** [The Persistent Grip of Social Class on College Admissions](https://www.nytimes.com/2021/05/26/upshot/college-admissions-essay-sat.html)
- What was the desing element you liked? 
- How do you think grammar of graphics was portrayed in the graph?



# Vectors in R {.smaller}
A vector is like a single column of data in R, it’s a list of values that are all the same type (like all numbers or all words).
Example: 
```{r}
#| output-location: column-fragment
# A vector of numbers
numbers <- c(1, 2, 3, 4, 5)

# A vector of words
fruits <- c("apple", "banana", "cherry")

# A vector of TRUE/FALSE values
logic <- c(TRUE, FALSE, TRUE)
```

# Data Frames in R {.smaller}
A data frame is like a table or spreadsheet — it has rows and columns, and each column can have its own type (numbers, words, etc.).
```{r}
#| output-location: column-fragment
# Create a data frame
students <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 21, 19),
  major = c("Math", "Physics", "Biology")
)

# Print the data frame
print(students)
```

# Data types in R {.smaller}
Data Types: A data type tells you what kind of data something is, like a number, a word (string), or a TRUE/FALSE value (logical).
```{R}
#| output-location: column-fragment
# Numeric (numbers)
num <- 42

# Character (text)
text <- "Hello, R!"

# Logical (TRUE/FALSE)
logic <- TRUE

# Print the values and their types
print(num)      # 42
print(text)     # "Hello, R!"
print(logic)    # TRUE

# Check the data type using class()
class(num)   # "numeric"
class(text)  # "character"
class(logic) # "logical"

```

## Code Example
Here’s an example of code and output using **ggplot2** in R.



## Two Columns {.smaller}

::: columns
::: column

Explanation

Columns are great for mixing code, plots, and text.\

- Left and right content\

- Works well for comparisons\

- Helps structure visual walkthroughs\
:::

::: {.column .fragment}
Example plot
:::
:::

# Wrap-Up

### Quarto slides make teaching reproducible

::: poll adds interactivity

{{< countdown >}} adds time control

YAML controls defaults (logos, themes, layout)
