---
title: Week 4
subtitle: Data-types 
format: stat198slides-revealjs
---

# Introduction

This is a sample deck for DataViz DeCal Slides. The features include:

1. Title slide with background image.
2. A bunch of yaml defaults.
3. Clickable logo in lower right that links to the previous page.
4. No logo or slide number on title slide.
5. Custom `:::poll` div for poll questions.
6. A [countdown timer](https://github.com/gadenbuie/countdown/tree/main/quarto) accessible by adding

   `{{< countdown "01:30" >}}`
   
   on a new line.


# Agenda

- Understand different types of variables and how they influence chart selection.  
- Understand how different purposes or ana;ytical tasks influence chart selection.
- Learn how to match data structures to appropriate visial encodings
- Acknowledge esistence of graphics all over the data analysis cycle, and how graphics can be broadly divided into "Exploration" and "Communication".

# GGPlot syntax intro



## {.smaller auto-animate=true}

::: {.columns}
::: {.column}
```{r}
#| label: "hi mom"
penguins |>
  select(bill_length_mm,
         flipper_length_mm,
         species)
```
:::
::: {.column .fragment}
```{r}
#| echo: false
#| fig-height: 9
penguins |>
  ggplot(aes(x = bill_length_mm,
             y = flipper_length_mm,
             color = species)) +
  geom_point() +
  theme_gray(base_size = 18)
```
:::
:::

::: notes
Draw arrows mapping variables to visual cues
:::

## {.smaller auto-animate=true}

::: {.columns}
::: {.column}
```{r}
#| eval: false
penguins |>
  ggplot(aes(x = bill_length_mm,
             y = flipper_length_mm,
             color = species)) +
  geom_point()
```
:::
::: {.column}
```{r}
#| echo: false
#| fig-height: 9
penguins |>
  ggplot(aes(x = bill_length_mm,
             y = flipper_length_mm,
             color = species)) +
  geom_point() +
  theme_gray(base_size = 18)
```
:::
:::


## `ggplot2()` syntax

. . .

`ggplot2()` builds a plot layer by layer, each one added on top of one another with `+` (not `|>`).

. . .

- `ggplot(df)` creates canvas
- `aes()` creates mappings, called inside `ggplot()` or a `geom()`
- `geom_()` puts down marks using declared geometry

## Layer by layer

```{r}
#| eval: false
penguins |>
  ggplot(aes(x = bill_length_mm,
             y = flipper_length_mm,
             color = species)) +
  geom_point() +
  theme_gray(base_size = 18)
```


## Common `aes()`

::: {.columns}
::: {.column}
- `x`
- `y`
- `color`
- `fill`
:::
::: {.column}
- `size`
- `shape`
- `alpha`
:::
:::

## Common `geom_()`

::: {.columns}
::: {.column}
- `geom_point()`
- `geom_bar() / geom_col()`
- `geom_line()`
:::
::: {.column}
- `geom_histogram()`
- `geom_boxplot()`
- `geom_violin()`
- `geom_density()`
:::
:::

## Mapping vs Setting

demo

<!-- 
## Two types of Viz

::: {.columns}
::: {.column}
### Explanatory Data Analysis
:::
::: {.column}
### Exploratory Data Analysis

:::
:::

::: notes
Add notes to each:

Explanatory:
- consumer: stakeholder
- goal: tell a specific story
- customized
- polished aesthetic

Exploratory:
- consumer: the analyst
- goal: uncover structure for further analysis
- follow templates
- prioritize quick and accurate interpretation
::: -->

# Discussion of Assigned Reading 

- Reading: The Persistent Grip of Social Class on College Admissions 
- Warm up question: 
- Follow up reading questions:
- In what ways do the graphs differ?

::: poll
What is your favorite color?

A. Red  
B. Green  
C. Blue  
:::

{{< countdown "01:30" >}}


## Code Example

Here’s an example of code and output using **ggplot2** in R.

```{r}
#| output-location: column-fragment
library(ggplot2)

ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(
    title = "Miles per Gallon vs. Weight",
    color = "Cylinders"
  )
```

## Two Columns {.smaller}

::: columns
::: column

Explanation

Columns are great for mixing code, plots, and text.\

- Left and right content\

- Works well for comparisons\

- Helps structure visual walkthroughs\
:::

::: {.column .fragment}
Example plot
:::
:::

# Wrap-Up

### Quarto slides make teaching reproducible

::: poll adds interactivity

{{< countdown >}} adds time control

YAML controls defaults (logos, themes, layout)
