---
title: Practice Quiz 1
format: typst
---

# Data Types & Chart Selection

## Types of Variables

 Assign the right number to the each sentence:

@.) Binary 
@.) Ordinal 
@.) Nominal 
@.) Continuous 
@.) Discrete 

___ Can take any countable value represented as an integer

___ Categories have no clear order, and are mutually exclusive

___ Only takes two possible values, and are typically represented as 1 or 0.

___ Have a clear ordering of the categories

___ Can take any measurable value within a given range, including fractions and decimals 


## Analytical Goals and Chart Selection

For the following datasets, **1)** come up with an analytical  goal, **2)** sate the variables you will use--and what type it is--to achieve it, and **3)** identify what chart will facilitate your observations.

```{r}
library(hce)

dataset1 = hce::ADLB

dataset1
```



@. You are given a dataset of laboratory measurements of kidney function over time. There are 8 variables: patient ID numbers, treatment values (1 = Active, 2 = Placebo), measurement value, measurement day in study

@. You are given a dataset of

@. You are given a dataset of

@. You are given a dataset of

@. You are given a dataset of


# Grammar of Graphics

## Understandig GG

@. GG allows creation of graphs to be _________, __________, and ________ 

@. What is aesthetic mapping?


@. Who formalized the grammar of graphics?

a) Clara Reckhorn
b) Leland Wilkinson
c) Stephen Few
d) Kieran Healy

@. Which of these  **isn't** part of the GG framework

a) Statistics
b) Theme
c) Analysis 
d) Coordinates

## Application

The

```{r echo=FALSE}
library(tidyverse)
library(ppgm)

fossil_dataset = ppgm::scel_fossils

fossil_dataset |>
  ggplot(aes(x =))

```


@. 

# Data Cleaning

## Open ended

@. Why is data cleaning important?

## Multiple Choice

A dataset contains the following column:

|Transfer Students|
|-----------------|
| Transfer|
| transfer|
| T |
| non transfer|
| NT |
| Non-transfer |

@. What is the main data cleaning issue here?

a) Missing values
b) Duplicate rows
c) Inconsistent categorical labels
d) Incorrect numerical values

@. Why is data cleaning necessary before running statistical analysis?

a) Because statistical models understand the meaning of messy data
b) Because computers automatically interpret inconsistent labels correctly
c) Because computers require structured, consistent inputs to produce valid results


@. If a company does not properly clean its sales data before building a predictive model, what is the most likely outcome?

a) The model will run faster
b) The model will automatically fix errors
c) The results may be biased or misleading
d) The model will become more accurate

# EDA

theoretical questions and multiple choice

# ggplot2

theoretical questions

what is ggplot2 based on
