---
title: Week 9 Activity 1
subtitle: Exploring Raster Data with terra
format: html
---

## Learning goals
::: {.column}
By the end of this activity, you will:

- load raster data using terra \ 

- inspect raster structure \

- visualize spatial grids \

- modify raster resolution \

- experiment with color scales \

- interpret raster meaning correctly \ 

:::
# Part 1 
```{r}
library(terra)
#example data included in the package
r <- rast(system.file("ex/elev.tif", package = "terra"))
r
```

Discuss with your partner \
## Question 1 

@. What type of spatial data is this? \
A) Vector
B) Raster

## Question 2
@. Each square cell represents:
A) A polygon boundary
B) A coordinate pair only
C) A value on a grid
D) A connected line segment

# Part 2 — Inspect Raster Properties

```{r}
nrow(r)
ncol(r)
res(r)
crs(r)
```

## Question 1 
@. What does resolution mean here?
A) Number of polygons
B) Distance between grid cells
C) Number of coordinate systems
D) Map projection accuracy

# Part 3 — Plot the Raster

```{r}
plot(r)
```

```{r}
#Now try a different color palette:
plot(r, col = terrain.colors(50))
```

```{r}
plot(r, col = heat.colors(50))
```

# Discuss
Which palette makes elevation differences easier to interpret? \ 

Why? \ 

# Part 4 — Change Raster Resolution

```{r}
r_lowres <- aggregate(r, fact = 5)

plot(r_lowres)
```

```{r}
#now compare
plot(r)
```

# Discuss
@. What changed?
A) Projection
B) Cell size
C) Coordinate system
D) Polygon boundaries


Which version shows more detail?

# Part 5- 
```{r}
ext_small <- ext(r, 6, 6.5, 49.5, 50)

r_crop <- crop(r, ext_small)

plot(r_crop)
```

@. Why might cropping raster data be useful?
Choose the best answer:
A) Reduces file size and improves focus
B) Changes projection accuracy
C) Converts raster to vector
D) Removes missing values automatically

# Part 6 — Compare Raster vs Choropleth Thinking (3 minutes)

Discuss with your partner:
@. When should you use raster instead of polygons? \ 

Choose the best answer: \

A) When mapping temperature or elevation
B) When mapping election results
C) When mapping state boundaries
D) When mapping city locations

# Final Reflection Question
In one sentence: \
@. What is the biggest difference between raster maps and choropleth maps?