---
title: Lecture 9
subtitle: Spatial Data Part 2 - Map Visualizations
format: stat198slides-revealjs
---

## Agenda

1.  Discussion of Reading

2. Maps Introduction

3. Activity 

4. Open notes quiz and Collaborative!!

## Final Project Elements {.smaller}

-   Final Project Proposal

    -   Type: Google doc pdf
    -   To: BCourses
    -   Due: 4/20, 11:59 PM

-   Presentation:

    -   Type: google slides
    -   Due: BCourses
    -   Due: 4/27, 11:59 PM

-   Coding Materials:

    -   Type: QMD file + csv of your data
    -   To: BCourses
    -   Due: 4/27, 11:59 PM

## ICEBREAKER

If your life had a dashboard, what’s one metric that would look concerning?

# Reading Discussion

Article: [These Maps Tell the Story of Two Americas: One Parched, One Soaked](https://www.nytimes.com/interactive/2021/08/24/climate/warmer-wetter-world.html)

# Why maps matter?
Imagine you have data for all 50 states.

A table looks like this:

- hard to see patterns
- hard to compare regions

But a map?

You instantly see:
- clusters
- outliers
- regional patterns

# Introduction to Maps

We already learned:

- Vector data → shapes (states, counties)
- Raster data → continuous surfaces
- Color → preattentive attention
- Gestalt → grouping patterns

Maps combine ALL of these ideas!

We are not just drawing shapes, we are attaching meaning to space.


# Before coding: what are we building?

We need two things:

1. Geometry (where things are)
2. Data (what we want to show)

Then we combine them.


# What makes NYT maps so powerful?
NYT maps are powerful because they:

::: incremental
- Show **clear geographic patterns**
- Use **simple color scales**
- Remove visual clutter
- Focus attention on one variable

# Small Example Maps Code (Loading Data)
```{r}
library(tidyverse)
library(maps)
library(viridis)

us_states <- map_data("state")
#Think: an empty coloring book.
head(us_states)
```

# Create Data for Map
```{r}
set.seed(123)

state_values <- tibble(
  region = unique(us_states$region),
  value = runif(length(unique(us_states$region)), 0, 100)
)
#Now every shape has a value attached.
map_df <- left_join(us_states, state_values, by = "region")
# We created fake data (incomes or rates)
# Then we JOIN it to map geometry
```

# First Map (Basic)
```{r}
ggplot(map_df, aes(long, lat, group = group, fill = value)) +
  geom_polygon(color = "white") + #draws states
  coord_fixed(1.3) +
  theme_void()
```


# Make it NYT-style (Basic)
```{r}
ggplot(map_df, aes(long, lat, group = group, fill = value)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_fixed(1.3) +
  scale_fill_viridis_c(option = "C") +
  theme_void()
```

# Introduction to Plotly 
Plotly is a package that turns static plots into: 

- interactive plots you can hover, zoom, and explore instead of just looking at a picture, you can interact with the data

Why do we use it?
Traditional plots (ggplot2):

- static image 
- no interaction 

Plotly:

- hover over points 
- zoom in/out 
- click and explore 

# Example
```{r}
library(plotly)
library(ggplot2)
#Do heavier cars use more gas?
p <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()

ggplotly(p)
```