---
title: "HW 4: Data Visualization"
subtitle: "Stat 198, Spring 2026"
fontsize: 12pt
urlcolor: blue
output: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Screen Time's Barchart Replication

The following table contains the data of the barchart (units: minutes).

| Days      | Social | Entertainment | Creativity | Other |
|:----------|:-------|:--------------|:-----------|:------|
| Sunday    | 107    | 29            | 1          | 101   |
| Monday    | 100    | 22            | 3          | 22    |
| Tuesday   | 46     | 32            | 8          | 69    |
| Wednesday | 88     | 14            | 1          | 44    |
| Thursday  | 117    | 0             | 1          | 72    |
| Friday    | 69     | 1             | 1          | 39    |
| Saturday  | 98     | 0             | 9          | 46    |

1.  Here's one way to obtain the input data table

```{r}
# input vectors
days = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
social = c(107, 100, 46, 88, 117, 69, 98)
entertainment = c(29, 22, 32, 14, 0, 1, 0)
creativity = c(1, 3, 8, 1, 1, 1, 9)
other = c(101, 22, 69, 44, 72, 39, 46)
```

2.  Assemble a data frame from the vectors above. Your column names should be: Day, Social, Entertainment, Creativity, Other. Store the result in a variable called `tbl`

*Hint: use data.frame() with the vectors as named columns.*

```{r}
# Assemble a data frame from the vectors above.


```

3.  Reshape tbl from wide format to long (tall) format so that ggplot can map category to fill color. The new column holding the category names should be called "Category", and the new column holding the minute values should be called "Time". Store the result back into `tbl`.

*Hint: use pivot_longer() — you want to pivot all columns except Day.*

```{r}
# Reshape tbl
```

```{.r}
# convert time from minutes to hours
tbl <- mutate(tbl, Time = Time / 60)

head(tbl)
```

4.  Compute total screen time per day (you'll use this to add average labels later). Store the result in a variable called `totals`.

*Hint: use group_by(Day) then summarize() to sum Time across all categories for each day.*

```{r}
# Compute total screen time per day

```

5.  Build a stacked bar chart using ggplot2.

*Hints:* - Use geom_col() with position = "stack"

-   Map Day to the x-axis, Time to y, and Category to fill

-   To control the order of days on the x-axis, convert Day to a factor with levels = days (the vector defined earlier) before plotting, or use scale_x_discrete()

-   Use scale_fill_manual(values = c(...)) to apply the color palette above

-   Store your ggplot object in a variable called `p` so you can add layers below

```{r}
# color palette (these names map to the four categories — assign them as you see fit)
orange1 = "#E59400"
light_gray1 = "#939393"
gray1 = "#4C4C4C"
blue1 = "#2D89E5"
light_blue1 = "#96C4F2"
green1 = "#00B300"

# Barchart

```

6.  Customize the theme to match the target chart.

*Hints:*

-   Start with theme_minimal() or theme_classic() as a base

-   Use theme() to adjust: axis labels, gridlines, legend position, etc.

-   To remove a theme element entirely, set it to element_blank()

-   To move the legend, use legend.position = inside the theme() call

-   Add axis and chart labels with labs(x = ..., y = ..., title = ..., fill = ...)

```{r}
# Customize the theme

```

7.  Add annotations to the chart (e.g. average daily screen time label).

*Hints:*

-   Use annotate("text", x = ..., y = ..., label = ...) for static text

-   Or use geom_text() / geom_label() with the `totals` data frame to place a label on each bar — set the `data` argument inside geom_text()

-   To display a horizontal reference line (e.g. average), use geom_hline()

-   Remember to print `p` at the end of this chunk (or just call p) so the final chart renders

```{r}
# Add annotations to the chart

```
