---
title: Dashboards
subtitle: "STAT 198"
format:
  stat198slides-revealjs:
    chalkboard: false
embed-resources: true
---

## Data Dashboards {auto-animate=true}

A *Data Dashboard* is an interactive display of selected statistics and plots to allow the user to quickly monitor, explore, and understand patterns in data.

Used for:

::: {.columns}
::: {.column}
1. **Real-time Monitoring**
   - [California Power Outages Map](https://www.arcgis.com/apps/dashboards/2f50a16613fa4f7f801b27a09a0be241){preview-link=true}
   - [Fremont Climate Action](https://www.fremont.gov/government/citywide-initiatives/data-dashboards/climate-action-dashboard){preview-link=true}
:::
::: {.column .fragment}
2. **Data Visualization**
   - [COVID 19 World Map Tracker](https://ourworldindata.org/explorers/covid?Metric=Excess+mortality+%28estimates%29&Interval=Cumulative&Relative+to+population=true&country=USA~BRA~JPN~DEU){preview-link=true}
   - [CalViz Degrees](https://calviz.berkeley.edu/t/OPAP/views/DegreeRecipientsByMajor/ExecutiveSummary)
:::
:::


## Quarto Dashboards

We can use the same `.qmd` files we have been using for activities (with small changes) to create clean, static dashboards!


# Demo


## Links:

Please open each of the following in a different browser tab.

*Documentation*: <https://quarto.org/docs/dashboards/>

*Dashboard Demo on BCourses (today's activity)*: <https://bcourses.berkeley.edu/courses/1553882/assignments/9070595>

> Download the qmd file and open it in your IDE.

::: notes
Click on link yourself and edit the dashboard on DataHub. First steps:

1. Render the doc as is and discuss how there are code cells that run and show but no output.
1. Have them add their name as author (so you can tell their submissions apart).
1. Run quarto publish to publish the HTML file.
1. Ask a student for their URL and project it to show that it is indeed publicly available.
2. Add format: dashboard.
:::

# Dashboard Anatomy

## Navbar

. . .

```markdown
--- 
title: "Palmer Penguins"
author: "Cobblepot Analytics"
format: 
  dashboard:
    logo: images/penguins.png
    nav-buttons: [linkedin, twitter, github]
---
```

<https://quarto.org/docs/reference/projects/websites.html#nav-items>{preview-link=true}


## Layout

- Define new row with `## Row`, new column with `## Column`
- Nest one inside the other using `###` for child (e.g. `### Column`)
- Defaults to row-by-row layout. Change by using

. . .

```markdown
---
format:
  dashboard:
    orientation: columns
---
```

- In general, each code cell creates one card
- Tune dimensions by adding e.g. `## Row {height=30%}`



## Layout by row {.smaller}

. . .

::: {.columns}
::: {.column}

````markdown
---
format: dashboard
---
    
## Row {height=70%}

```{{r}}
```

## Row {height=30%}

```{{r}}
```

```{{r}}
```
````
:::
::: {.column}
:::
:::

## Layout by column {.smaller}

. . .

::: {.columns}
::: {.column}
````markdown
---
format: 
  dashboard:
    orientation: columns
---
    
## Column {width=60%}

```{{r}}
```

## Column {width=40%}

```{{r}}
```

```{{r}}
```
````
:::
::: {.column}
:::
:::

## Your turn

1. Add four empty cards to your document by adding inserting four blank code cells at the bottom.
2. Arrange them in a 2 x 2 grid layout.

{{< countdown 2:30 >}}

# Data Display

## Plots {.smaller}

. . .

You can insert graphics from ggplot, base R, or an interactive library. It can be helpful to tweak `fig-width` and `fig-height`.

. . .

::: {.columns}
::: {.column}

````markdown
## Row {height="65%"}

```{{r}}
#| fig-width: 10
#| fig-height: 8
plot(cars)
```

## Row {height="35%"}
        
```{{r}}
#| fig-width: 5
#| fig-height: 4
plot(pressure)
```
    
```{{r}}
#| fig-width: 5
#| fig-height: 4
plot(airmiles)
```
````
:::
::: {.column}
:::
:::

## Tables {auto-animate=true}

1. Static table: `gt::gt`, `knitr::kable()`.
1. Interactive table: `DT::datatable()`


## `kable()` {.smaller}

```{r}
library(knitr)
kable(mtcars)
```


## `datatable()` {.smaller}
```{r}
library(DT)
datatable(mtcars)
```


## Value boxes

You can create a dynamic value box using the following.

````markdown
```{{r}}
#| component: valuebox
#| title: My big number
list(
  icon = "stopwatch",
  color = "primary",
  value = x
)
```
````

## Blank Cards

. . .

You can create a blank card to fill with anything (code, images, markdown) using:

```markdown
::: {.card}
Gapminder combines data from multiple sources into
unique coherent time-series that can’t be found
elsewhere. Learn more about the Gampminder dataset at
<https://www.gapminder.org/data/>.
:::
```

## Your Turn

Display the data on earthquakes in your dashboard according to the following layout.

{{< countdown 6:00 >}}


## Pages {.smaller}

Split your dashboard into multiple pages by adding level one headers with the page name. This example uses two pages, "Bills" and "Flippers", that will be linked in the navbar.

::: {.columns}
::: {.column}
````markdown
---
title: "Palmer Penguins"
format: dashboard
---
    
# Bills 

```{{r}}
```

# Flippers

## Column

```{{r}}
```

```{{r}}
```

## Column 

```{{r}}
```
````
:::
::: {.column}
:::
:::


# Submission

Once you've finished making these changes and testing out the dashboard formatting with different components of the demo, submit your `.qmd` file to the BCourses assignment. 

# Final Project Elements {.smaller}

-   Final Project Proposal

    -   Type: Google doc pdf
    -   To: BCourses
    -   Due: 4/20, 11:59 PM

-   Presentation:

    -   Type: google slides
    -   Due: BCourses
    -   Due: 4/27, 11:59 PM

-   Coding Materials:

    -   Type: QMD file + csv of your data
    -   To: BCourses
    -   Due: 4/27, 11:59 PM

## Project Work Time

- Questions on your project? Now's the time to ask us!

- Next week: 3-5 minutes per person to present your finished project to the class

- Great ideas so far! Excited to see the projects

```{r}
#| echo: false
countdown::countdown(35, top=0)
```

